---
title: "project 4 rough draft"
author: "Fernanda Montoya"
date: "2024-12-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
odd_dataset <- read.csv("/Users/fernandamontoya/Downloads/odd_combined_dataset.csv")

modified_dataset <- read.csv("/Users/fernandamontoya/Downloads/modified_dataset.csv")

even_dataset <- read.csv("/Users/fernandamontoya/Downloads/even_combined_dataset.csv")

combined_dataset <- read.csv("/Users/fernandamontoya/Downloads/combined_dataset.csv")
```

This chunk calls necessary libraries and forms the tsibble object that will be used for the TSLM models. Also included are lineplots for the outcomes of interest plotted by date

```{r, message = FALSE}
library(forecast)
library(fpp3)
library(tsibbledata)
library(tsibble)
library(fable)

# time series dataset able to be read in by TSLM using date(recorded in weeks) as the index
modified_ts <- tsibble::tsibble(
  date = as.Date(combined_dataset$date),
  NDVI = combined_dataset$NDVI,
  swe = combined_dataset$swe,
  EVI = combined_dataset$EVI,
  LAI = combined_dataset$LAI,
  prcp = combined_dataset$prcp,
  PET = combined_dataset$PET,
  tmax = combined_dataset$tmax,
  tmin = combined_dataset$tmin,
  vp = combined_dataset$vp,
  FPAR = combined_dataset$FPAR,
  ET = combined_dataset$ET,
  PLE = combined_dataset$PLE,
  dayl = combined_dataset$dayl,
  year = combined_dataset$year
)

# plot NDVI over time (higher values indicate healthier vegetation)
modified_ts %>% ggplot(aes(x = date, y = NDVI)) +
  geom_line(colour = "blue") +
  theme_bw() +
  labs(x = "Date", y = "NDVI", title = "Normalized Difference Vegetation Index Over Time")

# plot EVI over time (higher values indicate healthier vegetation)
modified_ts %>% ggplot(aes(x = date, y = EVI)) +
  geom_line(colour = "darkgreen") +
  theme_bw() +
  labs(x = "Date", y = "EVI", title = "Enhanced Vegetation Index Over Time")

# plot ET over time (lower values indicate reduced plant activity and heightened water stress)
modified_ts %>% ggplot(aes(x = date, y = ET)) +
  geom_line(colour = "cornflowerblue") +
  theme_bw() +
  labs(x = "Date", y = "ET (mm/day)", title = "Evapotranspiration Over Time")

# plot FPAR over time (lower values indicate a reduction in crop canopy density)
modified_ts %>% ggplot(aes(x = date, y = FPAR)) +
  geom_line(colour = "darkgoldenrod1") +
  theme_bw() +
  labs(x = "Date", y = "FPAR", title = "Fraction of Photosynthetically Active Radiation Over Time")

# plot LAI over time (lower values indicate diminished vegetation coverage or stunted growth)
modified_ts %>% ggplot(aes(x = date, y = LAI)) +
  geom_line(colour = "coral3") +
  theme_bw() +
  labs(x = "Date", y = "LAI", title = "Leaf Area Index Over Time")
```

This chunk performs the split of the training and test data and builds the model for one outcome, NDVI, as well as forcasting the performance of the model using the test data.

```{r}
# split of entire dataset to training and test (approx an 80-20 split)
train <- modified_ts %>% filter_index(~ "2019-12-27")
test <- modified_ts %>% filter_index("2020-01-03" ~.)

##NDVI
# time series linear model using crop health predictors and trend() to capture linear time trend
train_ndvi <- train %>% model(TSLM(NDVI ~ swe + prcp + tmax + tmin + vp + dayl + trend()))

# fit forecast with remaining test data
forecast_test_ndvi <- forecast(train_ndvi, new_data = test)

# compare predicted NDVI compared to entire time interval
forecast_test_ndvi %>% autoplot(modified_ts) + theme_bw() + labs(x = "Date", y = "NDVI", title = "NDVI Over Time")

# isolated prediction using the test data
forecast_test_ndvi %>% autoplot() + theme_bw() + labs(x = "Date", y = "NDVI", title = "NDVI Prediction")
```

This chunk repeats the above model and plots, but for the remaining outcomes of EVI, ET, FPAR, and LAI.

```{r}
## EVI
# time series linear model using crop health predictors and trend() to capture linear time trend
train_evi <- train %>% model(TSLM(EVI ~ swe + prcp + tmax + tmin + vp + dayl + trend()))

# fit forecast with remaining test data
forecast_test_evi <- forecast(train_evi, new_data = test)

# compare predicted NDVI compared to entire time interval
forecast_test_evi %>% autoplot(modified_ts, color = "darkgreen") + theme_bw() + labs(x = "Date", y = "EVI", title = "EVI Over Time")

# isolated prediction using the test data
forecast_test_evi %>% autoplot(color = "darkgreen") + theme_bw() + labs(x = "Date", y = "EVI", title = "EVI Prediction")

## ET
# time series linear model using crop health predictors and trend() to capture linear time trend
train_et <- train %>% model(TSLM(ET ~ swe + prcp + tmax + tmin + vp + dayl + trend()))

# fit forecast with remaining test data
forecast_test_et <- forecast(train_et, new_data = test)

# compare predicted NDVI compared to entire time interval
forecast_test_et %>% autoplot(modified_ts, color = "cornflowerblue") + theme_bw() + labs(x = "Date", y = "ET", title = "ET Over Time")

# isolated prediction using the test data
forecast_test_et %>% autoplot(color = "cornflowerblue") + theme_bw() + labs(x = "Date", y = "ET", title = "ET Prediction")

## FPAR
# time series linear model using crop health predictors and trend() to capture linear time trend
train_fpar <- train %>% model(TSLM(FPAR ~ swe + prcp + tmax + tmin + vp + dayl + trend()))

# fit forecast with remaining test data
forecast_test_fpar <- forecast(train_fpar, new_data = test)

# compare predicted NDVI compared to entire time interval
forecast_test_fpar %>% autoplot(modified_ts, color = "darkgoldenrod1") + theme_bw() + labs(x = "Date", y = "FPAR", title = "FPAR Over Time")

# isolated prediction using the test data
forecast_test_fpar %>% autoplot(color = "darkgoldenrod1") + theme_bw() + labs(x = "Date", y = "FPAR", title = "FPAR Prediction")

## LAI
# time series linear model using crop health predictors and trend() to capture linear time trend
train_lai <- train %>% model(TSLM(LAI ~ swe + prcp + tmax + tmin + vp + dayl + trend()))

# fit forecast with remaining test data
forecast_test_lai <- forecast(train_lai, new_data = test)

# compare predicted NDVI compared to entire time interval
forecast_test_lai %>% autoplot(modified_ts, color = "coral3") + theme_bw() + labs(x = "Date", y = "LAI", title = "LAI Over Time")

# isolated prediction using the test data
forecast_test_lai %>% autoplot(color = "coral3") + theme_bw() + labs(x = "Date", y = "LAI", title = "LAI Prediction")
```

This chunk looks at model performance on the training data by ploting the predicted values as well as the true values for each training model.

```{r, warning = FALSE}
## NDVI
p1 <- augment(train_ndvi) %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = NDVI), color = "dark gray", size = 1) +
    geom_line(aes(y = .fitted), color = "blue", size = 1) +
    theme_light() +
    labs(x = "Date", subtitle = "Time series")
p1

## EVI
p2 <- augment(train_evi) %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = EVI), color = "dark gray", size = 1) +
    geom_line(aes(y = .fitted), color = "darkgreen", size = 1) +
    theme_light() +
    labs(x = "Date", subtitle = "Time series")
p2

## ET
p3 <- augment(train_et) %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = ET), color = "dark gray", size = 1) +
    geom_line(aes(y = .fitted), color = "cornflowerblue", size = 1) +
    theme_light() +
    labs(x = "Date", subtitle = "Time series")
p3

## FPAR
p4 <- augment(train_fpar) %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = FPAR), color = "dark gray", size = 1) +
    geom_line(aes(y = .fitted), color = "darkgoldenrod1", size = 1) +
    theme_light() +
    labs(x = "Date", subtitle = "Time series")
p4

## LAI
p1 <- augment(train_lai) %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = LAI), color = "dark gray", size = 1) +
    geom_line(aes(y = .fitted), color = "coral3", size = 1) +
    theme_light() +
    labs(x = "Date", subtitle = "Time series")
p1
```





